Eagle: multi-locus association mapping on a genome-wide scale made routine

### 2019-07-24 BioinformaticsRevison/bioinformaticspaper.tex
Revesion 1

%%\documentclass{nature}
\documentclass{article}
\usepackage{xr}
\bibliographystyle{natbib}
\renewcommand{\baselinestretch}{1.4} 
\usepackage{graphicx}
\usepackage{bm}
\usepackage{lineno}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{authblk}

@@

\begin{abstract} \noindent
\setstretch{1.6}{
{\bf Motivation:} We present Eagle, a new method for multi-locus association mapping. The motivation for developing Eagle was to make multi-locus association mapping "easy" and the method-of-choice. Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. doesn't suffer from multiple testing issues c. is threshold free d. gives results that are immediately interpretable and e. has a computational footprint comparable to single-locus association mapping. \\
{\bf Results:} By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and multi-locus methods. We also analyse data from a published mouse study. Eagle found over 100\% more validated findings than the state-of-the-art single-locus method.  \\
{\bf Availability and Implementation:} Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users less familiar with R. It is freely available via the CRAN website  at https://cran.r-project.org.  \\
{\bf Contact:} andrew.george@csiro.au
}
\end{abstract}

\section{Introduction}
Over the past decade,  genome-wide association studies (GWASs) have changed considerably in both their analysis and design. Early studies
 followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification 
\citep{cardon2003population}. This prompted a shift towards family-based designs and score tests, such as the transmission/disequilibrium test (TDT)  and its variants \citep{spielman1996tdt}. Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification \citep{price2010new}. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine \citep{purcell2007plink, bradbury2007tassel}.

What has not changed is that it remains common practice to analyse genome-wide association study (GWAS) data on a locus-by-locus basis. This is despite there being several significant problems with analysing data in this way. 
First, the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. 
The identification of these regions from these analyses is not always straightforward. GWAS results are reported, typically, via Manhattan plots 
that plot the -log$_{10}$ of the $p$ value for each locus against the map position of the locus. The $p$ value is obtained by testing the statistical 
significance of a single nucleotide polymorphisms (SNP) when treated as an effect in an appropriate model. 
The location of peaks in this plot identify genomic 
regions of interest. Inferring the exact number of regions though can be difficult if the peaks are not well separated. Second, when multiple statistical tests are performed, the probability of wrongly accepting a result (type 1 error) is inflated. This is known as the multiple testing problem. 
Many different solutions have been offered \citep{storey2003statistical, li2005adjusting, de2005efficiency}. Yet, there is still no well accepted way of correcting for multiple testing in the context of genome-wide association mapping. Third, many of the traits whose genetic secrets we are trying to discover are complex. There will be multiple SNPs in linkage disequilibrium with genes that are influencing the trait. Yet, a locus-by-locus mapping approach only assesses the evidence for association between a single marker locus and trait.

It is somewhat surprising then that multi-locus association mapping methods haven't attracted more attention. Methods based on 
regularisation techniques, such as ridge regression \citep{shen2013novel}  and lasso \citep{rakitsch2013lasso}, measure all locus-trait associations simultaneously. 
Here, multiple testing is not an issue. These techniques though are computationally demanding. Also, their results can be difficult to interpret. The strength of association is not measured by a $p$ value but by the size of the regression coefficient for the SNP in the model. More recently, associations have started to be mapped with random forests \citep{szymczak2016r2vim}. Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings. A multi-locus method that does show promise is the multiple-locus linear mixed model method \citep{segura2012efficient}. The best multi-locus model is built with simple forward selection. Results are immediately interpretable but here, computation becomes challenging for large datasets.

@@

The data were obtained from a large genome-wide association study that was performed in outbred mice \citep{nicod2016genome}. 
Phenotypic and genotypic data were available on 1,887 adult mice. 
The phenotypic data included raw and adjusted (for fixed effects) measurements from 200 behavioural, tissue, and physiological traits.  
Of these traits, 
43 yielded SNP-trait associations that could be corroborated through other independent published work. It was these 
43 traits that were the focus of our real data analyses. As in the original study  \citep{nicod2016genome}, our analyses 
were based on the adjusted traits.
Genotypic data were available on 359, 559 (353,697 autosomal) SNPs in the 
form of marker dosages (expected allele counts that ranged from zero to one). All missing data had been imputed. 
We converted the dosages into discrete genotypes 
by clustering around 0, 0.5, and 1, corresponding to SNP genotypes AA, AB, and BB, respectively. 
We focused our analyses on the autosomal SNPs.

@@

Eagle is a method for multi-locus association mapping on a genome-wide scale. It is based on linear mixed models. It differs from most other single- and multi-locus association mapping methods. Eagle treats association mapping as a model selection instead of variable selection problem. Consequently,  we do not have to contend with multiple testing issues or having to construct significance thresholds.  
Eagle also 
reports as its findings only those SNPs that are in strongest linkage disequilibrium, and hence closest to the genes influencing a trait. 
The methodological foundation for Eagle comes from a whole-genome linkage analysis method that was developed for mapping 
quantitative trait loci in experimental crosses \citep{verbyla2007analysis}.

@@

\section{Discussion/Conclusion}
Eagle is a new linear mixed model based method (and R package) for multi-locus association mapping. It advances the state of association mapping in several ways. 
First, its computational footprint is much smaller than other multi-locus implementations. Eagle makes multi-locus analysis 
practical, even when the datasets are large. Second, the results from
 Eagle are immediately interpretable. They are the set of SNPs in strongest association with the trait where 
each SNP identifies a separate genomic region of interest. Third, it treats association mapping as a model selection problem, avoiding 
multiple testing issues and the need for significance thresholds. 
 As we saw in the simulation study, Eagle has considerably higher power than single-locus methods but is comparable in run time.
Also, when analysing the mouse data, Eagle found more than double the SNP-trait associations than 
with single-locus association mapping, the method of choice. Furthermore, these extra findings were all true.

>>>>

%%\documentclass{nature}
\documentclass{article}
\usepackage{xr}
\bibliographystyle{natbib}
\renewcommand{\baselinestretch}{1.4} 
\usepackage{graphicx}
\usepackage{bm}
\usepackage{lineno}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{authblk}
\usepackage{xcolor}

@@

\begin{abstract} \noindent
\setstretch{1.6}{
{\bf Motivation:} We present Eagle, a new method for multi-locus association mapping. The motivation for developing Eagle was to make multi-locus association mapping "easy" and the method-of-choice. Eagle's strengths are that it a. is considerably more powerful than single-locus association mapping b. does not suffer from multiple testing issues c. gives results that are immediately interpretable and d. has a computational footprint comparable to single-locus association mapping. \\
{\bf Results:} By conducting a large simulation study, we will show that Eagle finds true and avoids false SNP-trait associations better than competing single- and multi-locus methods. We also analyse data from a published mouse study. Eagle found over 100\% more validated findings than the state-of-the-art single-locus method.  \\
{\bf Availability and Implementation:} Eagle has been implemented as an R package, with a web-based Graphical User Interface (GUI) for users less familiar with R. It is freely available via the CRAN website  at https://cran.r-project.org.  \\
{\bf Contact:} andrew.george@csiro.au
}
\end{abstract}

\section{Introduction}
Over the past decade,  genome-wide association studies (GWASs) have changed considerably in both their analysis and design. Early studies
 followed a case-control design. Association mapping methods were no more complicated than contingency table tests or simple 
linear regression. These designs though had a tendency to yield spurious findings if there was unrecognised population stratification 
\citep{cardon2003population}. This prompted a shift towards family-based designs and score tests, such as the transmission/disequilibrium test (TDT)  and its variants \citep{spielman1996tdt}. Today, instead of by design, it is through statistical modelling that we account for the effects of population stratification \citep{price2010new}. This has meant that data can be collected from general populations, even if these populations are highly structured. Analysis via sophisticated association mapping methods, such as linear mixed model based approaches,  is now almost routine  \citep{yu2006unified,zhao2007arabidopsis}.

What has not changed is that it remains common practice to analyse genome-wide association study (GWAS) data on a locus-by-locus basis. This is despite there being several significant problems with analysing data in this way. 
First, \textcolor{red}{ for each SNP, a hypothesis test is performed. The null hypothesis is that  there is no association between the SNP and trait. 
The alternative is that the SNP is in association with the trait. It is straight forward to guard against wrongly rejecting the null 
hypothesis (or making a type 1 error) if only a single hypothesis test is being performed. However, the analysis of GWAS data 
with locus-by-locus methods necessitates  conducting a large number of correlated hypothesis tests, simultaneously. 
 This leads to an increased risk of type 1 errors.  To deal with this challenge, many different solutions have been offered 
 \citep{storey2003statistical, li2005adjusting, de2005efficiency}}.  Second, 
the aim of association mapping is to identify regions of the genome that house genes that are influencing a trait. 
The identification of these regions from these analyses is not always straightforward. GWAS results are reported, typically, via Manhattan plots 
that plot the -log$_{10}$ of the $p$ value for each locus against the map position of the locus. The $p$ value is obtained from the hypothesis test. 
The location of peaks in this plot identify genomic 
regions of interest. Inferring the exact number of regions though can be difficult if the peaks are not well separated. 
 Third, many of the traits whose genetic secrets we are trying to discover are complex. There will be multiple SNPs in linkage disequilibrium with genes that are influencing the trait. Yet, a locus-by-locus mapping approach only assesses the evidence for association between a single marker locus and trait.

It is somewhat surprising then that multi-locus association mapping methods haven't attracted more attention. Methods based on 
regularisation techniques, such as ridge regression \citep{shen2013novel}  and lasso \citep{rakitsch2013lasso}, measure all locus-trait associations simultaneously. These techniques though are computationally demanding. \textcolor{red}{Also, the strength of association is not measured by a $p$ value but by the size of the regression coefficient for the SNP in the model. Further processing is required before the results can be interpreted \citep{cho2010joint, rakitsch2013lasso}}.
More recently, associations have started to be mapped with random forests \citep{szymczak2016r2vim}. Similar to regularisation techniques though, it is not clear how to infer genomic regions of interest from their findings. A multi-locus method that does show promise is the multiple-locus linear mixed model method \citep{segura2012efficient}. The best multi-locus model is built with forward and backward stepwise selection. 
Results are immediately interpretable \textcolor{red}{in that the SNP closest to the genes underlying the trait are identified} 
but computation does become challenging for large datasets.

@@

The data were obtained from a large genome-wide association study that was performed in outbred mice \citep{nicod2016genome}. 
Phenotypic and genotypic data were available on 1,887 adult mice. 
The phenotypic data included raw and adjusted (for fixed effects) measurements from 200 behavioural, tissue, and physiological traits.  
Of these traits, 
43 yielded SNP-trait associations that could be corroborated through other independent published work. It was these 
43 traits that were the focus of our real data analyses. As in the original study  \citep{nicod2016genome}, our analyses 
were based on the adjusted traits.
Genotypic data were available on \textcolor{red}{359,559} (353,697 autosomal) SNPs in the 
form of marker dosages (expected allele counts that ranged from zero to one). All missing data had been imputed. 
We converted the dosages into discrete genotypes 
by clustering around 0, 0.5, and 1, corresponding to SNP genotypes AA, AB, and BB, respectively. 
We focused our analyses on the autosomal SNPs.

@@

Eagle is a method for multi-locus association mapping on a genome-wide scale. It is based on linear mixed models. It differs from most other single- and multi-locus association mapping methods in that Eagle treats association mapping as a model selection problem (ball2001bayesian,broman2002model,yi2005bayesian). 
The "best" model is found via forward selection. It makes use of a modified form of the Bayesian information criterion, BIC, for model selection.

Eagle does not suffer from multiple testing issues. A "best" model is built iteratively. At each iteration, a hypothesis test is performed. 
Only a small number of iterations are needed in building the "best" model.  In contrast,  for single-locus methods, multiple testing is an 
issue because each SNP is assessed separately, culminating in a large number of hypothesis tests being performed. 
Eagle also 
reports as its findings only those SNPs that are in strongest linkage disequilibrium with  the genes influencing a trait. 
The methodological foundation for Eagle comes from a whole-genome linkage analysis method that was developed for mapping 
quantitative trait loci in experimental crosses \citep{verbyla2007analysis}.

@@

\section{Discussion/Conclusion}
Eagle is a new linear mixed model based method (and R package) for multi-locus association mapping. It advances the state of association mapping in several ways. 
First, its computational footprint is much smaller than other multi-locus implementations. Eagle makes multi-locus analysis 
practical, even when the datasets are large. Second, the results from
 Eagle are immediately interpretable. They are the set of SNPs in strongest association with the trait where 
each SNP identifies a separate genomic region of interest. Third, it treats association mapping as a model selection problem, avoiding 
 the need for significance thresholds. 
 As we saw in the simulation study, Eagle has considerably higher power than single-locus methods but is comparable in run time.
Also, when analysing the mouse data, Eagle found more than double the SNP-trait associations than 
with single-locus association mapping, the method of choice. Furthermore, these extra findings were all true.
